---
title: "Analysis of Wildlife Strikes in Commercial Aviation"
author: "Kenish Raghu"
date: "10-23-2024"
output:
  html_document: default
  pdf_document: default
subtitle: Practicum I CS5200
---

```{r loadingLib, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
library(DBI)
library(RMySQL)
library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)

```


# Setting Up Database Connection Credentials
```{r db_credentials, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
host <- "kendb2.cx0g202cysew.us-east-1.rds.amazonaws.com"  # AWS RDS endpoint
port <- 3306  # MySQL default port
dbname <- "kendb2"  # Name of my database
user <- "admin"  # my MySQL username
password <- "kenish123"  # my MySQL password
```

# Establishing Connection to Database
```{r establishingConnection, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
con <- dbConnect(RMySQL::MySQL(),
                 host = host,
                 port = port,
                 dbname = dbname,
                 user = user,
                 password = password)

# Checking if the connection is successful
if(!is.null(con)) {
  print("Connection successful!")
} else {
  print("Connection failed.")
}
```

## Loading csv data into a dataframe
```{r loadingCSV, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Loading CSV data into a dataframe
bds.raw <- read.csv("BirdStrikesData-V4-SuF24.csv", stringsAsFactors = FALSE)
bds.raw[is.na(bds.raw)] <- "unknown"
bds.raw[bds.raw == ""] <- "unknown"

# Proof of unknown being inserted
bds.raw[bds.raw$iid == 208561, ]

```

## Dropping tables ( if they exist )
```{r droppingTables, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Dropping Tables if they already exist
dbExecute(con, "DROP TABLE IF EXISTS wildlife_strikes")
dbExecute(con, "DROP TABLE IF EXISTS conditions")
dbExecute(con, "DROP TABLE IF EXISTS flights")
dbExecute(con, "DROP TABLE IF EXISTS airports")
dbExecute(con, "DROP TABLE IF EXISTS wildlife_sizes")

```

## Table creation 
```{r creatingTable, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Creating the airports table
create_airports_table <- "
CREATE TABLE airports (
  aid INT PRIMARY KEY AUTO_INCREMENT,
  airportName TEXT NOT NULL,
  airportState TEXT NOT NULL,
  airportCode VARCHAR(3) DEFAULT 'ZZZ'
);"

dbExecute(con, create_airports_table)


# Creating the flights table
create_flights_table <- "
CREATE TABLE flights (
  fid INT PRIMARY KEY,
  date DATE,
  originAirport INT,
  airlineName TEXT,
  aircraftType TEXT,
  isHeavy INT,
  FOREIGN KEY (originAirport) REFERENCES airports (aid)
);"

dbExecute(con, create_flights_table)


# Creating the conditions lookup table
create_conditions_table <- "
CREATE TABLE conditions (
  cid INT PRIMARY KEY AUTO_INCREMENT,
  sky_condition TEXT NOT NULL,
  explanation TEXT
);"

dbExecute(con, create_conditions_table)


# Creating the wildlife_sizes lookup table
create_wildlife_size_table <- "
CREATE TABLE wildlife_sizes (
  wlsizeid INT PRIMARY KEY AUTO_INCREMENT,
  description TEXT NOT NULL
);"

dbExecute(con, create_wildlife_size_table)


# Creating the wildlife_strikes table with foreign key to wildlife_sizes
create_wildlife_strikes_table <- "
CREATE TABLE wildlife_strikes (
  iid INT PRIMARY KEY AUTO_INCREMENT,
  fid INT,
  wlsize INT,
  impact TEXT NOT NULL,
  altitude INT CHECK (altitude >= 0),
  conditions INT,
  FOREIGN KEY (fid) REFERENCES flights (fid),
  FOREIGN KEY (wlsize) REFERENCES wildlife_sizes (wlsizeid),
  FOREIGN KEY (conditions) REFERENCES conditions (cid)
);"

dbExecute(con, create_wildlife_strikes_table)

```

##  Verifying Table Definitions are correct
```{r verificationOfTables, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Defining the list of tables to check for existence 
tables_to_check <- c("airports", "flights", "conditions", "wildlife_sizes", "wildlife_strikes")

# Function to check and display only existing tables
get_existing_tables <- function(con, tables) {
  # Querying to get the list of tables in the database
  existing_tables <- dbGetQuery(con, "SHOW TABLES;")
  
  # Extracting table names from the query result
  existing_table_names <- existing_tables[[1]]
  
  # Finding tables that exist in both the database and the list
  existing_tables_in_check <- intersect(existing_table_names, tables)
  
  # Returning only existing table names as a data frame
  data.frame(Table_Name = existing_tables_in_check, stringsAsFactors = FALSE)
}

# Running the check and display results
existing_tables_df <- get_existing_tables(con, tables_to_check)
print(existing_tables_df)

```


## Inserting data into 'Airports'
```{r airportInsertion, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Function to create and populate the airports table dataframe
create_airports_df <- function(bds.raw) {
  
  unique_airports <- unique(bds.raw[, c("dep_airport", "origin_state")])
  
  airports_df <- data.frame(
    aid = seq_len(nrow(unique_airports)),
    airportName = unique_airports$dep_airport,
    airportState = unique_airports$origin_state,
    airportCode = "ZZZ",
    stringsAsFactors = FALSE
  )
  
  return(airports_df)
}

airports_df <- create_airports_df(bds.raw)
dbWriteTable(con, name = "airports", value = airports_df, row.names = FALSE, append = TRUE)

```


## Inserting data into 'Conditions'
```{r conditionInsertion, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Function to create and populate the conditions table dataframe
create_conditions_df <- function(bds.raw) {
  unique_conditions <- unique(bds.raw$sky_conditions)
  
  conditions_df <- data.frame(
    cid = seq_along(unique_conditions),
    sky_condition = unique_conditions,
    explanation = "",
    stringsAsFactors = FALSE
  )
  
  return(conditions_df)
}

conditions_df <- create_conditions_df(bds.raw)
dbWriteTable(con, name = "conditions", value = conditions_df, row.names = FALSE, append = TRUE)

```



## Inserting data into 'Wildlife_Sizes'
```{r wildlifeSizeInsertion, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Function to create and populate the wildlife sizes table dataframe
create_wildlife_size_df <- function(bds.raw) {
  unique_sizes <- unique(bds.raw$wildlife_size)
  
  wildlife_size_df <- data.frame(
    wlsizeid = seq_along(unique_sizes),
    description = unique_sizes,
    stringsAsFactors = FALSE
  )
  
  return(wildlife_size_df)
}

wildlife_size_df <- create_wildlife_size_df(bds.raw)
dbWriteTable(con, name = "wildlife_sizes", value = wildlife_size_df, row.names = FALSE, append = TRUE)

```


##  Inserting data into 'Flights'
```{r flightInsertion, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Function to create and populate the flights table dataframe
process_flights_df <- function(bds.raw, airports_df) {
  
  bds.raw$airport_key <- paste(bds.raw$dep_airport, bds.raw$origin_state, sep = "_")
  airports_df$airport_key <- paste(airports_df$airportName, airports_df$airportState, sep = "_")
  
  match_indices <- match(bds.raw$airport_key, airports_df$airport_key)
  bds.raw$originAirport <- airports_df$aid[match_indices]
  
 # bds.raw$airline[is.na(bds.raw$airline) | bds.raw$airline == ""] <- "Unknown"
  bds.raw$aircraftType <- bds.raw$aircraft
 # bds.raw$isHeavy <- ifelse(bds.raw$heavy_flag == "Yes", TRUE, FALSE)
  bds.raw$isHeavy <- ifelse(
  bds.raw$heavy_flag == "Yes", 
  1, 
  ifelse(bds.raw$heavy_flag == "unknown" | bds.raw$heavy_flag == "", NA, 0)
)
  
  bds.raw$date <- as.Date(bds.raw$flight_date, format = "%m/%d/%y")
  
  flights_df <- data.frame(
    fid = bds.raw$iid,
    date = bds.raw$date,
    originAirport = bds.raw$originAirport,
    airlineName = bds.raw$airline,
    aircraftType = bds.raw$aircraftType,
    isHeavy = as.integer(bds.raw$isHeavy),
    stringsAsFactors = FALSE
  )
  
  return(flights_df)
}

flights_df <- process_flights_df(bds.raw, airports_df)
dbWriteTable(con, name = "flights", value = flights_df, row.names = FALSE, append = TRUE)

```

## Inserting data into 'Wildlife_strikes'
```{r wildlifeStrikeInsertion, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Function to create and populate the wildlife_strikes table dataframe
process_incidents_df <- function(bds.raw, conditions_df, wildlife_size_df, flights_df) {

  # Assign fid values from flights_df to bds.raw
  bds.raw$fid <- flights_df$fid

  # Map conditions to get cid values from conditions_df
  match_conditions <- match(bds.raw$sky_conditions, conditions_df$sky_condition)
  bds.raw$conditions <- conditions_df$cid[match_conditions]

  # Map wildlife size to get wlsizeid from wildlife_size_df
  match_sizes <- match(bds.raw$wildlife_size, wildlife_size_df$description)
  bds.raw$wlsize <- wildlife_size_df$wlsizeid[match_sizes]

  # Handle altitude, replacing invalid values with 0
  bds.raw$altitude_ft[is.na(bds.raw$altitude_ft) | bds.raw$altitude_ft == "" | bds.raw$altitude_ft < 0] <- 0
  bds.raw$altitude_ft <- gsub(",", "", bds.raw$altitude_ft)  # Remove commas if present
  bds.raw$altitude <- as.integer(bds.raw$altitude_ft)

  # Assign unique incident IDs (iid)
  bds.raw$iid <- seq_len(nrow(bds.raw))

  # Create incidents_df for wildlife_strikes table insertion
  incidents_df <- data.frame(
    iid = bds.raw$iid,
    fid = bds.raw$fid,
    wlsize = bds.raw$wlsize,
    impact = bds.raw$impact,
    altitude = bds.raw$altitude,
    conditions = bds.raw$conditions,
    stringsAsFactors = FALSE
  )

  return(incidents_df)
}

# Run the process and insert into the wildlife_strikes table
incidents_df <- process_incidents_df(bds.raw, conditions_df, wildlife_size_df, flights_df)
dbWriteTable(con, name = "wildlife_strikes", value = incidents_df, row.names = FALSE, append = TRUE)

```


```{r displayingTablesForVerification, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Display the first few rows of the wildlife.strikes table
wildlife_strikes <- dbGetQuery(con, "SELECT * FROM wildlife_strikes LIMIT 10;")
print(wildlife_strikes)

# Display the first few rows of the flights table
flights <- dbGetQuery(con, "SELECT * FROM flights LIMIT 10;")
print(flights)

# Display the first few rows of the airports table
airports <- dbGetQuery(con, "SELECT * FROM airports LIMIT 10;")
print(airports)

# Display the first few rows of the wildlife_size table
airports <- dbGetQuery(con, "SELECT * FROM wildlife_sizes LIMIT 10;")
print(wildlife_size_df)

```

## Top States with Strikes
```{r topFiveStates, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
# SQL query to find the top 5 states with the greatest number of wildlife strike incidents
top_states_query <- "
  SELECT a.airportState, COUNT(*) AS num_incidents
  FROM wildlife_strikes ws
  JOIN flights f ON ws.fid = f.fid
  JOIN airports a ON f.originAirport = a.aid
  GROUP BY a.airportState
  ORDER BY num_incidents DESC
  LIMIT 5;
"

# Execute the query and store the results
top_states <- dbGetQuery(con, top_states_query)

# Display the results in a colorful, formatted table
top_states %>%
  kable("html", col.names = c("State", "Number of Incidents"),caption = "<b>Top 5 States by Wildlife Strike Incidents</b>") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"), position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#FF5722") %>%     # Header styling
  column_spec(1, bold = TRUE, background = "#FFCC80", color = "#BF360C") %>% # State column
  column_spec(2, background = "#FFEB3B", color = "#F57F17")                  # Number of Incidents column
```

## Analysis by Airport
```{r airportsWithBelowAverageIncidents, echo=FALSE, eval = TRUE, warning = FALSE}
# SQL query to find airports with below average wildlife strike incidents
analysis_airport_query <- "
  SELECT airportName, COUNT(ws.iid) AS num_incidents
  FROM wildlife_strikes ws
  JOIN flights f ON ws.fid = f.fid
  JOIN airports a ON f.originAirport = a.aid
  GROUP BY airportName
  HAVING num_incidents < (
      SELECT AVG(num_incidents)
      FROM (
          SELECT COUNT(ws.iid) AS num_incidents
          FROM wildlife_strikes ws
          JOIN flights f ON ws.fid = f.fid
          JOIN airports a ON f.originAirport = a.aid
          GROUP BY airportName
      ) AS avg_incidents
  )
  ORDER BY num_incidents DESC
  LIMIT 5;
"

# Execute the query and store the results
below_avg_airports <- dbGetQuery(con, analysis_airport_query)


# Display the results in a colorful, formatted table
below_avg_airports %>%
  kable("html", col.names = c("Airport Name", "Number of Incidents"), caption = "<b>Airports with below Avg. Wildlife Strike Incidents</b>") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"), position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4CAF50") %>%     # Header styling
  column_spec(1, bold = TRUE, background = "#AED581", color = "#1B5E20") %>% # Airport Name column
  column_spec(2, background = "#FFF176", color = "#F57F17")                  # Number of Incidents column
```

## Analysis by Year
```{r numberOfTypeOfIncidentsTotalPerYear, echo=FALSE, eval = TRUE, warning = FALSE}
# SQL query to find the number of small, medium, and large wildlife strikes per year
analysis_year_query <- "
  SELECT YEAR(f.date) AS year, 
         SUM(CASE WHEN ws.wlsize = 2 THEN 1 ELSE 0 END) AS small_strikes,
         SUM(CASE WHEN ws.wlsize = 1 THEN 1 ELSE 0 END) AS medium_strikes,
         SUM(CASE WHEN ws.wlsize = 3 THEN 1 ELSE 0 END) AS large_strikes
  FROM wildlife_strikes ws
  JOIN flights f ON ws.fid = f.fid
  GROUP BY year
  ORDER BY year;
"

# Execute the query and store the results in a dataframe
results <- dbGetQuery(con, analysis_year_query)
# Remove rows with NA values
year_analysis <- na.omit(results)

# Display the results in a colorful, formatted table
year_analysis %>%
  kable("html", col.names = c("Year", "Small Strikes", "Medium Strikes", "Large Strikes"), caption = "<b>Wildlife Strikes by Size</b>") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"), position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4CAF50") %>%   # Header styling
  column_spec(1, bold = TRUE, background = "#FFDDC1", color = "#A52A2A") %>% # Year column
  column_spec(2, background = "#D4E157", color = "#33691E") %>%            # Small Strikes column
  column_spec(3, background = "#FFD54F", color = "#FF6F00") %>%            # Medium Strikes column
  column_spec(4, background = "#FF8A80", color = "#B71C1C")                # Large Strikes column
```

## Trend by Year
```{r trendFollowedByYear, echo=FALSE, eval = TRUE, warning = FALSE}
# Createing a new column for total strikes by summing small, medium, and large strikes
year_analysis$total_strikes <- year_analysis$small_strikes + 
                               year_analysis$medium_strikes + 
                               year_analysis$large_strikes

y_range <- range(year_analysis$total_strikes)

# Plotting the trend of total strikes by year
plot(year_analysis$year, year_analysis$total_strikes, type = "o", col = "blue", 
     xlab = "Year", 
     ylab = "Total Number of Strikes", 
     main = "Wildlife Strikes Trend by Year",
     ylim = c(y_range[1], y_range[2] * 1.2),
     pch = 16,               # Point type
     lwd = 2)                # Line width

# Adding grid lines for better readability
grid()

# Adding text labels for each point
text(year_analysis$year, year_analysis$total_strikes, labels = year_analysis$total_strikes, pos = 3, cex = 0.8)

```

## Stored Procedure Creation, Execution & Verification
```{r storedProcedure, echo=FALSE, eval = TRUE, warning = FALSE}
# Creating the update_log table if it does not exist
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS update_log (
  log_id INT PRIMARY KEY AUTO_INCREMENT,
  modification_type VARCHAR(50),
  table_name VARCHAR(50),
  updated_at DATETIME,
  original_values TEXT
);
")

# Dropping PROCEDURE IF EXISTS
dbExecute(con, "DROP PROCEDURE IF EXISTS update_wildlife_proced4;")


# 2. Create the stored procedure for updating a strike
dbExecute(con, "
  CREATE PROCEDURE update_wildlife_proced4 (
  IN strike_id INT,
  IN new_wlsize INT,
  IN new_impact TEXT,
  IN new_altitude INT,
  IN new_conditions INT
)
BEGIN
  DECLARE original_wlsize INT;
  DECLARE original_impact TEXT;
  DECLARE original_altitude INT;
  DECLARE original_conditions INT;

  -- Retrieve original values for logging
  SELECT wlsize, impact, altitude, conditions 
  INTO original_wlsize, original_impact, original_altitude, original_conditions
  FROM wildlife_strikes 
  WHERE iid = strike_id;

  -- Log the update
  INSERT INTO update_log (modification_type, table_name, updated_at, original_values)
  VALUES ('UPDATE', 'wildlife_strikes', NOW(), 
          CONCAT('IID: ', strike_id, ', Original wlsize: ', original_wlsize, 
                 ', Original impact: ', original_impact, 
                 ', Original altitude: ', original_altitude, 
                 ', Original conditions: ', original_conditions));

  -- Update the incident
  UPDATE wildlife_strikes
  SET wlsize = new_wlsize, impact = new_impact, altitude = new_altitude, conditions = new_conditions
  WHERE iid = strike_id;
END ;
")

# Call the stored procedure to update a wildlife strike
# Replace 202152 with an existing ID from your wildlife_strikes table
dbExecute(con, "CALL update_wildlife_proce(202152, 3, 'Wing Damage', 1200, 2);")

# Check if the strike was updated
updated_strike <- dbGetQuery(con, "SELECT * FROM wildlife_strikes WHERE fid = 202152;")
print("Updated Strike:")
print(updated_strike)

# Check the update_log to see if the log entry was created
log_result <- dbGetQuery(con, "SELECT * FROM update_log ORDER BY updated_at DESC;")
print("Update Log Entries:")
print(log_result)

# Print the result as a colorful table
log_result %>%
  kable("html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4CAF50") %>%  # Header row
  column_spec(1, bold = TRUE, color = "white", background = "#FF5733") %>% # Column 1 styling
  column_spec(2, color = "#2E86C1") %>% # Column 2 styling
  column_spec(3, background = "#AED6F1") %>% # Column 3 styling
  column_spec(4, italic = TRUE)  # Column 4 styling

```
## Disconnect from Database
```{r disconnectDatabase, eval = T, warning = F}
# Disconnect from the database
dbDisconnect(con)

```


