---
title: "Analysis of Wildlife Strikes in Commercial Aviation"
subtitle: "Practicum I CS5200"
author: "Ashwin Reddy"
email: "reddy.ash@northeastern.edu"
date: "29th October 2024"
semester: "Fall 2024"
output: html_document
---

<br>
<br>

## 1. Loading Required Libraries
```{r loadingLibraries, eval = T, warning = F}
# Load the necessary libraries
library(DBI)
library(RMySQL)
library(kableExtra)

```

<br>
<br>

## 2. Connecting to AWS MySQL Database
```{r connectingToAWSMySQL, eval = T, warning = F}
host <- "assignment4.c982ko2u6peq.us-east-1.rds.amazonaws.com"  # AWS RDS endpoint
port <- 3306  # MySQL default port
dbname <- "assignment4"  
user <- "admin"  
password <- "admin123"  

# Establish the connection
con <- dbConnect(RMySQL::MySQL(),
                 host = host,
                 port = port,
                 dbname = dbname,
                 user = user,
                 password = password)

# Check if the connection is successful
if(!is.null(con)) {
  print("Connection successful!")
} else {
  print("Connection failed.")
}

```

<br>
<br>

## 3. Dropping tables ( if they exist )
```{r dropTables, eval = T, warning = F}
# Drop Tables if they already exist
dbExecute(con, "DROP TABLE IF EXISTS wildlife_strikes")
dbExecute(con, "DROP TABLE IF EXISTS conditions")
dbExecute(con, "DROP TABLE IF EXISTS flights")
dbExecute(con, "DROP TABLE IF EXISTS airports")
dbExecute(con, "DROP TABLE IF EXISTS wildlife_sizes")

```

<br>
<br>

## 4. Table creation 
```{r createTables, eval = T, warning = F}
# Creating the airports table
create_airports_table <- "
CREATE TABLE airports (
  aid INT PRIMARY KEY AUTO_INCREMENT,
  airportName TEXT NOT NULL,
  airportState TEXT NOT NULL,
  airportCode VARCHAR(3) DEFAULT 'ZZZ'
);"

dbExecute(con, create_airports_table)


# Creating the flights table
create_flights_table <- "
CREATE TABLE flights (
  fid INT PRIMARY KEY,
  date DATE,
  originAirport INT,
  airlineName TEXT,
  aircraftType TEXT,
  isHeavy INT,
  FOREIGN KEY (originAirport) REFERENCES airports (aid)
);"

dbExecute(con, create_flights_table)


# Creating the conditions lookup table
create_conditions_table <- "
CREATE TABLE conditions (
  cid INT PRIMARY KEY AUTO_INCREMENT,
  sky_condition TEXT NOT NULL,
  explanation TEXT
);"

dbExecute(con, create_conditions_table)


# Creating the wildlife_sizes lookup table
create_wildlife_size_table <- "
CREATE TABLE wildlife_sizes (
  wlsizeid INT PRIMARY KEY AUTO_INCREMENT,
  description TEXT NOT NULL
);"

dbExecute(con, create_wildlife_size_table)


# Creating the wildlife_strikes table with foreign key to wildlife_sizes
create_wildlife_strikes_table <- "
CREATE TABLE wildlife_strikes (
  iid INT PRIMARY KEY AUTO_INCREMENT,
  fid INT,
  wlsize INT,
  impact TEXT NOT NULL,
  altitude INT CHECK (altitude >= 0),
  conditions INT,
  FOREIGN KEY (fid) REFERENCES flights (fid),
  FOREIGN KEY (wlsize) REFERENCES wildlife_sizes (wlsizeid),
  FOREIGN KEY (conditions) REFERENCES conditions (cid)
);"

dbExecute(con, create_wildlife_strikes_table)

```

<br> 
<br>

## 5. Verifying Table Definitions are correct
```{r verifyingTableDefinitions, eval = F, warning = F}
# Verify Table Definitions and Column Structure
verify_table_structure <- function(con, table_name, expected_columns) {
  # Retrieve column information from the database
  column_info <- dbGetQuery(con, paste0("DESCRIBE ", table_name, ";"))
  # Check if all expected columns are in the table
  actual_columns <- column_info$Field
  missing_columns <- setdiff(expected_columns, actual_columns)
  if (length(missing_columns) == 0) {
    message(paste("Table '", table_name, "' structure is correct."))
  } else {
    message(paste("Table '", table_name, "' is missing columns:", paste(missing_columns, collapse = ", ")))
  }
}

# Expected columns for flights table
expected_flights_columns <- c("fid", "date", "originAirport", "airlineName", "aircraftType", "isHeavy")
verify_table_structure(con, "flights", expected_flights_columns)

# Expected columns for airports table
expected_airports_columns <- c("aid", "airportName", "airportState", "airportCode")
verify_table_structure(con, "airports", expected_airports_columns)

# Expected columns for conditions table
expected_conditions_columns <- c("cid", "sky_condition", "explanation")
verify_table_structure(con, "conditions", expected_conditions_columns)

# Expected columns for wildlife_sizes table
expected_wildlife_sizes_columns <- c("wlsizeid", "description")
verify_table_structure(con, "wildlife_sizes", expected_wildlife_sizes_columns)

# Expected columns for wildlife_strikes table
expected_wildlife_strikes_columns <- c("iid", "fid", "wlsize", "impact", "altitude", "conditions")
verify_table_structure(con, "wildlife_strikes", expected_wildlife_strikes_columns)

```

<br> 
<br>

## 6. Loading csv data into a dataframe
```{r loadCSV, eval = T, warning = F}
# Load CSV data into a dataframe
bds.raw <- read.csv("BirdStrikesData.csv", stringsAsFactors = FALSE)
bds.raw[is.na(bds.raw)] <- "unknown"
bds.raw[bds.raw == ""] <- "unknown"

# Proof of unknown being inserted
bds.raw[bds.raw$iid == 208561, ]

```

<br>
<br>

## 7. Inserting data into 'Airports'
```{r dataInsertionAirports, eval = T, warning = F}
# Function to create and populate the airports table dataframe
create_airports_df <- function(bds.raw) {
  #bds.raw$dep_airport[is.na(bds.raw$dep_airport) | bds.raw$dep_airport == ""] <- "Unknown"
  #bds.raw$origin_state[is.na(bds.raw$origin_state) | bds.raw$origin_state == ""] <- "Unknown"
  
  unique_airports <- unique(bds.raw[, c("dep_airport", "origin_state")])
  
  airports_df <- data.frame(
    aid = seq_len(nrow(unique_airports)),
    airportName = unique_airports$dep_airport,
    airportState = unique_airports$origin_state,
    airportCode = "ZZZ",
    stringsAsFactors = FALSE
  )
  
  return(airports_df)
}

airports_df <- create_airports_df(bds.raw)
dbWriteTable(con, name = "airports", value = airports_df, row.names = FALSE, append = TRUE)

```

<br>
<br>

## 8. Inserting data into 'Conditions'
```{r dataInsertionConditions, eval = T, warning = F}
# Function to create and populate the conditions table dataframe
create_conditions_df <- function(bds.raw) {
 # bds.raw$sky_conditions[is.na(bds.raw$sky_conditions) | bds.raw$sky_conditions == ""] <- "Unknown"
  unique_conditions <- unique(bds.raw$sky_conditions)
  
  conditions_df <- data.frame(
    cid = seq_along(unique_conditions),
    sky_condition = unique_conditions,
    explanation = "",
    stringsAsFactors = FALSE
  )
  
  return(conditions_df)
}

conditions_df <- create_conditions_df(bds.raw)
dbWriteTable(con, name = "conditions", value = conditions_df, row.names = FALSE, append = TRUE)

```

<br>
<br>

## 9. Inserting data into 'Wildlife_Sizes'
```{r dataInsertionWildlifeSizes, eval = T, warning = F}
# Function to create and populate the wildlife sizes table dataframe
create_wildlife_size_df <- function(bds.raw) {
 # bds.raw$wildlife_size[is.na(bds.raw$wildlife_size) | bds.raw$wildlife_size == ""] <- "Unknown"
  unique_sizes <- unique(bds.raw$wildlife_size)
  
  wildlife_size_df <- data.frame(
    wlsizeid = seq_along(unique_sizes),
    description = unique_sizes,
    stringsAsFactors = FALSE
  )
  
  return(wildlife_size_df)
}

wildlife_size_df <- create_wildlife_size_df(bds.raw)
dbWriteTable(con, name = "wildlife_sizes", value = wildlife_size_df, row.names = FALSE, append = TRUE)

```

<br>
<br>

## 10. Inserting data into 'Flights'
```{r dataInsertionFlights, eval = T, warning = F}
# Function to create and populate the flights table dataframe
process_flights_df <- function(bds.raw, airports_df) {
 # bds.raw$dep_airport[is.na(bds.raw$dep_airport) | bds.raw$dep_airport == ""] <- "Unknown"
 # bds.raw$origin_state[is.na(bds.raw$origin_state) | bds.raw$origin_state == ""] <- "Unknown"
  
  bds.raw$airport_key <- paste(bds.raw$dep_airport, bds.raw$origin_state, sep = "_")
  airports_df$airport_key <- paste(airports_df$airportName, airports_df$airportState, sep = "_")
  
  match_indices <- match(bds.raw$airport_key, airports_df$airport_key)
  bds.raw$originAirport <- airports_df$aid[match_indices]
  
 # bds.raw$airline[is.na(bds.raw$airline) | bds.raw$airline == ""] <- "Unknown"
  bds.raw$aircraftType <- bds.raw$aircraft
 # bds.raw$isHeavy <- ifelse(bds.raw$heavy_flag == "Yes", TRUE, FALSE)
  bds.raw$isHeavy <- ifelse(
  bds.raw$heavy_flag == "Yes", 
  1, 
  ifelse(bds.raw$heavy_flag == "unknown" | bds.raw$heavy_flag == "", NA, 0)
)
  
  bds.raw$date <- as.Date(bds.raw$flight_date, format = "%m/%d/%y")
  
  flights_df <- data.frame(
    fid = bds.raw$iid,
    date = bds.raw$date,
    originAirport = bds.raw$originAirport,
    airlineName = bds.raw$airline,
    aircraftType = bds.raw$aircraftType,
    isHeavy = as.integer(bds.raw$isHeavy),
    stringsAsFactors = FALSE
  )
  
  return(flights_df)
}

flights_df <- process_flights_df(bds.raw, airports_df)
dbWriteTable(con, name = "flights", value = flights_df, row.names = FALSE, append = TRUE)

```

<br>
<br>

## 11. Inserting data into 'Wildlife_strikes'
```{r dataInsertionWildlifeStrikes, eval = T, warning = F}
# Function to create and populate the wildlife_strikes table dataframe
process_incidents_df <- function(bds.raw, conditions_df, wildlife_size_df, flights_df) {

  # Step 2: Assign fid values from flights_df to bds.raw
  bds.raw$fid <- flights_df$fid

  # Step 3: Clean and match values for conditions and wildlife size
 # bds.raw$sky_conditions[is.na(bds.raw$sky_conditions) | bds.raw$sky_conditions == ""] <- "Unknown"
 # bds.raw$wildlife_size[is.na(bds.raw$wildlife_size) | bds.raw$wildlife_size == ""] <- "Unknown"

  # Map conditions to get cid values from conditions_df
  match_conditions <- match(bds.raw$sky_conditions, conditions_df$sky_condition)
  bds.raw$conditions <- conditions_df$cid[match_conditions]

  # Map wildlife size to get wlsizeid from wildlife_size_df
  match_sizes <- match(bds.raw$wildlife_size, wildlife_size_df$description)
  bds.raw$wlsize <- wildlife_size_df$wlsizeid[match_sizes]

  # Step 4: Handle altitude, replacing invalid values with 0
  bds.raw$altitude_ft[is.na(bds.raw$altitude_ft) | bds.raw$altitude_ft == "" | bds.raw$altitude_ft < 0] <- 0
  bds.raw$altitude_ft <- gsub(",", "", bds.raw$altitude_ft)  # Remove commas if present
  bds.raw$altitude <- as.integer(bds.raw$altitude_ft)

  # Step 5: Clean up impact field
 # bds.raw$impact[is.na(bds.raw$impact) | bds.raw$impact == ""] <- "Unknown"

  # Step 6: Assign unique incident IDs (iid)
  bds.raw$iid <- seq_len(nrow(bds.raw))

  # Step 7: Create incidents_df for wildlife_strikes table insertion
  incidents_df <- data.frame(
    iid = bds.raw$iid,
    fid = bds.raw$fid,
    wlsize = bds.raw$wlsize,
    impact = bds.raw$impact,
    altitude = bds.raw$altitude,
    conditions = bds.raw$conditions,
    stringsAsFactors = FALSE
  )

  return(incidents_df)
}

# Run the process and insert into the wildlife_strikes table
incidents_df <- process_incidents_df(bds.raw, conditions_df, wildlife_size_df, flights_df)
dbWriteTable(con, name = "wildlife_strikes", value = incidents_df, row.names = FALSE, append = TRUE)

```

<br>
<br>

## 12. Verifying 'Airports' table data insertion
```{r verifyingAirportsInsertion, eval = T, warning = F}
# Query to retrieve all records from the airports table
airports_data <- dbGetQuery(con, "SELECT * FROM airports ;")
print(airports_data)

```

<br>
<br>

## 13. Verifying 'Conditions' table data insertion
```{r verifyingConditionsInsertion, eval = T, warning = F}
# Query to retrieve all records from the conditions table
conditions_data <- dbGetQuery(con, "SELECT * FROM conditions LIMIT 25;")
print(conditions_data)

```

<br>
<br>

## 14. Verifying 'Flights' table data insertion
```{r verifyingFlightsInsertion, eval = T, warning = F}
# Query to retrieve all records from the flights table
flights_data <- dbGetQuery(con, "SELECT * FROM flights where isHeavy is NULL;")
print(flights_data)

```

<br>
<br>

## 15. Verifying 'Wildlife_Sizes' table data insertion
```{r verifyingWildlifeSizesInsertion, eval = T, warning = F}
# Query to retrieve all records from the wildlife_sizes table
wildlife_size_data <- dbGetQuery(con, "SELECT * FROM wildlife_sizes LIMIT 25;")
print(wildlife_size_data)

```

<br>
<br>

## 16. Verifying 'WildlifeStrikes' table data insertion
```{r verifyingWildlifeStrikesInsertion, eval = T, warning = F}
# Query to retrieve all records from the wildlife_strikes table
strikes_data <- dbGetQuery(con, "SELECT * FROM wildlife_strikes limit 25 ")
print(strikes_data)

```

<br>
<br>

## 17. Verifying Key constraints in 'Flights'
```{r verifyingKeyConstraintsFlights, eval = T, warning = F}
# Attempt to insert a duplicate primary key into flights table
tryCatch({
  dbExecute(con, "INSERT INTO flights (fid, flight_date, originAirport, airlineName, aircraftType, isHeavy) 
                  VALUES (1195, '2024-10-25', 1, 'Sample Airline', 'Sample Aircraft', 0);")
}, error = function(e) {
  print("Primary key constraint is working: Duplicate insert blocked.")

})

```

<br>
<br>

## 18. Verifying Key constraints in 'WildlifeStrikes'
```{r verifyingKeyConstraintsWildlifeStrikes, eval = T, warning = F}
# Attempt to insert a wildlife strike with a non-existent flight ID
tryCatch({
  dbExecute(con, "INSERT INTO wildlife_strikes (fid, wlsize, impact, altitude, conditions) 
                  VALUES (9999, 'Medium', 'Wing Damage', 500, 1);")
}, error = function(e) {
  print("Foreign key constraint is working: Invalid reference blocked.")
})

```

<br>
<br>

## 19. Verifying Key constraints in 'Airports'
```{r verifyingKeyConstraintsAirports, eval = T, warning = F}
# Attempt to delete an airport with associated flights
tryCatch({
  dbExecute(con, "DELETE FROM airports WHERE aid = 1;")
}, error = function(e) {
  print("Foreign key constraint on delete is working: Dependent row deletion blocked.")
})

```

<br>
<br>

## 20. Top States with Strikes
```{r queryTop5StatesWithHighestIncidents, echo=F, eval = T, warning = F}
# SQL query to find the top 5 states with the greatest number of wildlife strike incidents
top_states_query <- "
  SELECT a.airportState, COUNT(*) AS num_incidents
  FROM wildlife_strikes ws
  JOIN flights f ON ws.fid = f.fid
  JOIN airports a ON f.originAirport = a.aid
  GROUP BY a.airportState
  ORDER BY num_incidents DESC
  LIMIT 5;
"

# Execute the query and store the results
top_states <- dbGetQuery(con, top_states_query)

# Display the results in a nicer format using kable
top_states %>%
  kable("html") %>%
  kable_styling(full_width = F, position = "left")

```

<br>
<br>

## 21. Analysis by Airport
```{r queryAirportsWithBelowAverageIncidents, echo=F, eval = T, warning = F}
# SQL query to find airports with below average wildlife strike incidents
analysis_airport_query <- "
  SELECT airportName, COUNT(ws.iid) AS num_incidents
  FROM wildlife_strikes ws
  JOIN flights f ON ws.fid = f.fid
  JOIN airports a ON f.originAirport = a.aid
  GROUP BY airportName
  HAVING num_incidents < (
      SELECT AVG(num_incidents)
      FROM (
          SELECT COUNT(ws.iid) AS num_incidents
          FROM wildlife_strikes ws
          JOIN flights f ON ws.fid = f.fid
          JOIN airports a ON f.originAirport = a.aid
          GROUP BY airportName
      ) AS avg_incidents
  )
  ORDER BY num_incidents DESC
  LIMIT 5;
"

# Execute the query and store the results
below_avg_airports <- dbGetQuery(con, analysis_airport_query)

# Display the results in a nicer format
below_avg_airports %>%
  kable("html") %>%
  kable_styling(full_width = F, position = "left")

```

<br>
<br>

## 22. Analysis by Year
```{r queryNumberOfTypeOfIncidentsTotallyPerYear, echo=F, eval = T, warning = F}
# SQL query to find the number of small, medium, and large wildlife strikes per year
## Analysis by Year
# SQL query to find the number of small, medium, and large wildlife strikes per year
analysis_year_query <- "
  SELECT YEAR(f.date) AS year, 
         SUM(CASE WHEN ws.wlsize = 2 THEN 1 ELSE 0 END) AS small_strikes,
         SUM(CASE WHEN ws.wlsize = 1 THEN 1 ELSE 0 END) AS medium_strikes,
         SUM(CASE WHEN ws.wlsize = 3 THEN 1 ELSE 0 END) AS large_strikes
  FROM wildlife_strikes ws
  JOIN flights f ON ws.fid = f.fid
  GROUP BY year
  ORDER BY year;
"

# Execute the query and store the results in a dataframe
year_analysis <- dbGetQuery(con, analysis_year_query)

# Display the results in a formatted table
year_analysis %>%
  kable("html", col.names = c("Year", "Small Strikes", "Medium Strikes", "Large Strikes")) %>%
  kable_styling(full_width = F, position = "left")

```

<br>
<br>

## 23. Trend by Year
```{r queryTrendByYear, echo=F, eval = T, warning = F}
# Create a new column for total strikes by summing small, medium, and large strikes
year_analysis$total_strikes <- year_analysis$small_strikes + 
                               year_analysis$medium_strikes + 
                               year_analysis$large_strikes

y_range <- range(year_analysis$total_strikes)

# Plotting the trend of total strikes by year
plot(year_analysis$year, year_analysis$total_strikes, type = "o", col = "blue", 
     xlab = "Year", 
     ylab = "Total Number of Strikes", 
     main = "Wildlife Strikes Trend by Year",
     ylim = c(y_range[1], y_range[2] * 1.2),
     pch = 16,               # Point type
     lwd = 2)                # Line width

# Adding grid lines for better readability
grid()

# Adding text labels for each point
text(year_analysis$year, year_analysis$total_strikes, labels = year_analysis$total_strikes, pos = 3, cex = 0.8)

```

<br>
<br>

## 24. Stored Procedure Creation, Execution & Verification
```{r storedProcedure, eval = T, warning = F}
# 1. Create the update_log table if it does not exist
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS update_log (
  log_id INT PRIMARY KEY AUTO_INCREMENT,
  modification_type VARCHAR(50),
  table_name VARCHAR(50),
  updated_at DATETIME,
  original_values TEXT
);
")

# 2. Create the stored procedure for updating a strike
dbExecute(con, "
  CREATE PROCEDURE update_wildlife_proce (
  IN strike_id INT,
  IN new_wlsize INT,
  IN new_impact TEXT,
  IN new_altitude INT,
  IN new_conditions INT
)
BEGIN
  DECLARE original_wlsize INT;
  DECLARE original_impact TEXT;
  DECLARE original_altitude INT;
  DECLARE original_conditions INT;

  -- Retrieve original values for logging
  SELECT wlsize, impact, altitude, conditions 
  INTO original_wlsize, original_impact, original_altitude, original_conditions
  FROM wildlife_strikes 
  WHERE iid = strike_id;

  -- Log the update
  INSERT INTO update_log (modification_type, table_name, updated_at, original_values)
  VALUES ('UPDATE', 'wildlife_strikes', NOW(), 
          CONCAT('IID: ', strike_id, ', Original wlsize: ', original_wlsize, 
                 ', Original impact: ', original_impact, 
                 ', Original altitude: ', original_altitude, 
                 ', Original conditions: ', original_conditions));

  -- Update the incident
  UPDATE wildlife_strikes
  SET wlsize = new_wlsize, impact = new_impact, altitude = new_altitude, conditions = new_conditions
  WHERE iid = strike_id;
END ;
")

# 3. Step 1: Call the stored procedure to update a wildlife strike
# Replace 202152 with an existing ID from your wildlife_strikes table
dbExecute(con, "CALL update_wildlife_proce(202152, 3, 'Wing Damage', 1200, 2);")

# Step 2: Check if the strike was updated
updated_strike <- dbGetQuery(con, "SELECT * FROM wildlife_strikes WHERE fid = 202152;")
print("Updated Strike:")
print(updated_strike)

# Step 3: Check the update_log to see if the log entry was created
log_entries <- dbGetQuery(con, "SELECT * FROM update_log ORDER BY updated_at DESC;")
print("Update Log Entries:")
print(log_entries)

```

<br>
<br>

## 25. Disconnect from Database
```{r disconnectDatabase, eval = T, warning = F}
# Disconnect from the database
dbDisconnect(con)

```






